---
title: "CSSS 594, Homework 1"
author: "Neal Marquez"
date: "Due: October 10, 2017"
output:
    html_document:
    theme: cosmo
highlight: haddock
---

# Question 1: 

Suppose you are interested in studying the effects of physical disability on
depression. Over the last 5 years, you have collected data on adults ages 18 to
55 who have become physically disabled at some point during that time, based on 
a registry of diagnosed patients at a hospital. You were able to interview each
individual soon after their disability was diagnosed (during the first month) &
follow-up two times, between 6 and 8 months and between 12 and 16 months after
the diagnosis. The same procedure was conducted with a control group of the same
sample size. Each time, depression was assessed using a standardized scale.

 a. What is the outcome? How is it measured?
 b. What is the time metric? Where is zero? How many measurement occasions
    per individual? Time-structured or -unstructured schedule?
 c. Would an analysis that used the described data to address the research 
    question, about the association between disability and depression, be 
    considered a longitudinal data analysis? Why or why not?
    
## Part a

The outcome that was assessed was depression, which was measured using a 
standardized scale.

## Part b

The time schedule is sessions after disability diagnosis. Time point zero is the
session that occurs within the first monthe of diagnosis. There are a total of
three time points of data collected per individual. This apperas to be a time
structured data collection process although there seems to be a level of
flexibility within the three time periods for which data is collected.

## Part c

This analysis could be seen as a logitudinal data analysis because we have
repeated measures of the same individuals over time and can look at person level
differences as well as group level differences.


# Question 2.

From Singer and Willett (2003), Section 1.1.3:

''Many psychiatrists find that short-term anxiety-provoking psychotherapy (STAPP) can ameliorate psychological distress. A methodological strength of the associated literature is its consistent use of a well-developed instrument: the __Symptom Check-List__<sup>a<sub>1</sub></sup> (SCL-90), developed by __Derogatis (1994)__<sup>a<sub>2</sub></sup>. A methodological weakness is its reliance on two-wave designs: one wave of data pre-treatment and a second wave post-treatment. Researchers conclude that the treatment is effective when the decrease in SCL-90 scores among STAPP patients is lower than the decrease among individuals in a comparison group.

Svartberg, Seltzer, Stiles and Khoo (1995) adopted a different approach to studying STAPP's efficacy. Instead of collecting just two waves of data, the researchers examined ``the course, rate and correlates of symptom improvement as measured with the SCL-90 during and after STAPP'' (p. 242). A sample of 15 patients received approximately 20 weekly STAPP sessions. During the study, each patient completed the SCL-90 up to __seven times__<sup>b<sub>1</sub></sup>: __once or twice at referral (before therapy began)__,<sup>b<sub>2</sub></sup> __once at mid-therapy, once at termination and three times after therapy ended (after 6, 12 and 24 months)__<sup>b<sub>3</sub></sup>. Suspecting that STAPP's effectiveness would vary with the patients' abilities to control their emotional and motivational impulses (known as \emph{ego rigidity}), two independent psychiatrists reviewed the patients' intake files and assigned  __ego rigidity ratings__<sup>b<sub>4</sub></sup>.''

# Question 3
## Part a

```{r}
data <- read.csv("./tolerance1_pp.csv")
```

## Part b

```{r}
summary(data)
```

## Part c

```{r}
library(ggplot2)
data$Male <- as.factor(data$male)
ggplot(data, aes(x=exposure, y=tolerance, shape=Male)) + 
    geom_point()
```

## Part d

```{r}
maledata <- subset(data, male == 1)
femaledata <- subset(data, male == 0)
```

## Part e

```{r}
malemintime <- min(maledata$time)
malerangetol <- round(range(maledata$tolerance) ,2)
femalemintime <- min(femaledata$time)
femalerangetol <- round(range(femaledata$tolerance) ,2)
```

For males, the minimum value of time is `r malemintime`. The range of tolerance is (`r malerangetol`).

For females, the minimum value of time is `r femalemintime`. The range of tolerance is (`r femalerangetol`).

## Part f

```{r}
lm1 <- lm(tolerance ~ time + male + exposure, data=data)
```

## Part g

```{r}
library(pander)
pander(summary(lm1))
```